О множествах точек на плоскости с целочисленными расстояниями

### 2019-06-19 pps-linear-bound/pps-linear-lower-bound.tex
Final fixes after self-proofreading

\section{Introduction}
Let $\mathbb{N}$ be the set of all positive integers and let $|M_1 M_2|$ denote the Euclidean distance
between points $M_1$ and $M_2$ in a finite-dimensional space $\mathbb{R}^m$
(and, more generally, let $|\Delta|$ denote the length of line segment $\Delta$).
An \textit{integral point set} in $m$-dimensional Euclidean space is a point set $M$ such that all the distances between the
points of $M$ are integers and $M$ is not situated on an $(m-1)$-dimensional hyperplane.
Erd\"os and Anning proved~\cite{anning1945integral,erdos1945integral} that every integral point set consists of a finite number of points.
Taking the fact into account, we will denote the set of all integral point sets of $n$ points in $m$-dimensional Euclidean space with
$\mathfrak{M}(m,n)$ (using the notation in~\cite{our-vmmsh-2018}),
and let the symbol $\mathfrak{M}(m,\mathbb{N})$ stand for the set of all integral point sets in $m$-dimensional Euclidean space.
The symbol $\# M$ will be used for cardinality of $M$, i.e. the quantity of points in $M$ in our case.

@@

In this section,
we improve a better lower bound for minimum diameter of planar integral point sets by
involving Point Packing in a Square problem.
Below we introduce the problem, basic notions and results.

@@

\begin{definition}
	\textit{Cross} for points $M_1$ and $M_2$, denoted by $cr(M_1,M_2)$, is union of two straight lines:
	the line which the points $M_1$ and $M_2$ are situated on,
	and the bisector of line segment $M_1 M_2$.
\end{definition}

@@

Performing some simple manipulations with the upper bound in Theorem~\ref{thm:varphi_k_bounds},
one can make the following proposition.
\begin{proposition}
	\label{varphi_n_where_n_geq_21492}
	For $n \geq 21492$ we have
	\begin{equation}
		\varphi_n \leq \varphi_{n-1} \leq \frac{\beta}{\sqrt{n-2}}
		,
	\end{equation}
	where
	\begin{equation}
		\beta = \frac{1}{\sqrt{21490}} + \sqrt{ \frac{2}{\sqrt{3}} + \frac{1}{21490} } < 1.07464
		.
	\end{equation}
\end{proposition}

Now we need to estimate the quanity of points in an intersection of an integral point set with a line segment.
There is the following estimate for minimum diameter of planar integral point sets with many collinear points.
\begin{theorem}~\cite[Theorem 4]{kurz2008minimum}
	For $\delta > 0$, $\varepsilon > 0$, and $P\in\mathfrak{M}(2,n)$ with
	at least $n^\delta$ collinear points there exists a $n_0 (\varepsilon)$
	such that for all $n \geq n_0 (\varepsilon)$ we have
	\begin{equation}
		\operatorname{diam} P \geq n^{\frac{\delta}{4 \log 2(1+\varepsilon)}\log \log n}
		.
	\end{equation}
\end{theorem}
However, the estimate is rather unsuitable for our needs,
as it does not provide the values of all the constants.
To obtain the desired estimate, we will now prove a generalization of~\cite[lemma 3]{our-vmmsh-2018}.

@@

\begin{proposition}
	\label{obs:rho_curves}
	If points $M_1,M_2,M_3,M_4$ are situated on a straight line,
	and equality $|M_1 M_2| = |M_3 M_4|$ holds but line segments $M_1 M_2$ and $M_3 M_4$ do not coincide,
	then for a fixed $k$ the $\rho(k,M_1 M_2)$-curve and $\rho(k,M_3 M_4)$-curve
	do not intersect.
\end{proposition}

@@

\begin{proof}
	Let $f(k)$ denote the mininal length of a line segment
	that shares $k$ points with and integral point set.
	We observe that $f(k) > f (k-1)$.
	Due to Lemma~\ref{lem:line_segment_with n_squared_plus_one_points},
	$f(n^2+1) \geq \frac{2}{3}n^3+\frac{1}{2}n^2-\frac{1}{6}n$.

@@

\begin{theorem}
	\label{thm:main_estimate}
	If $n\geq 4$, then $d(2,n) \geq \gamma (n - 2)$,
	where
	\begin{multline}
		0.46530... =
		\frac{3^{1/4}}{2^{3/2}} >
		\\ >
		\gamma = \frac{\sqrt{16 {{\left( \sqrt{\frac{2}{\sqrt{3}}+\frac{1}{21490}}+\frac{1}{\sqrt{21490}}\right) }^{2}}+\frac{14791716}{{{\left( 2593 \sqrt{647}-5823\right) }^{2}}}}-\frac{3846}{2593 \sqrt{647}-5823}}{8 {{\left( \sqrt{\frac{2}{\sqrt{3}}+\frac{1}{21490}}+\frac{1}{\sqrt{21490}}\right) }^{2}}}
		>\\
		> 0.45557
		> \frac{5}{11}
		.
	\end{multline}
\end{theorem}

@@

Facher sets are the simplest ones of planar integral point sets.
It is shown that for $9\leq n \leq 122$ all the optimal sets are facher~\cite{kurz2008minimum}.
For every cardinality $n$ and every characteristic $q$
there exists a facher set $M\in\mathfrak{M}(2,n)$ with characteristic $q$~\cite[Theorem 5]{our-vmmsh-2018}.
In~\cite{antonov2008maximal}, the facher sets of characteristic 1 were investigated; they were called \textit{semi-crabs}.

@@

	We will now define the coordinates of the points as following:
	\begin{equation}
		M_{J\pm} =\left(\pm\frac{b_J}{2}, 0\right)
		,
		~~
		N   =\left(0, \frac{\sqrt{a}}{2}\right)
		.
	\end{equation}
	Then the distances are:
	\begin{multline}
		|N M_{J\pm}|
		=
		\left(\frac{b_J^2}{4} + \frac{a}{4}\right)^{1/2}
		=
		\frac{1}{2}\left(\left(\frac{c_J - a/c_J}{2}\right)^2 + a\right)^{1/2}
		=
		\\=
		\frac{1}{2}\left( \left(\frac{c_J}{2}\right)^2 - \frac{a}{2} + \left(\frac{a}{c_j}\right)^2 + a\right)^{1/2}
		=
		\frac{1}{2}\left( \left(\frac{c_J}{2}\right)^2 + \frac{a}{2} + \left(\frac{a}{c_j}\right)^2    \right)^{1/2}
		=
		\\=
		\frac{1}{2}\left(\left(\frac{c_J + a/c_J}{2}\right)^2\right)^{1/2}
		=
		\frac{1}{2}\left(\frac{c_J + a/c_J}{2}\right)
		=
		\frac{g_J}{2}
		\in\mathbb{N}
		,
	\end{multline}
	\begin{equation}
		|M_{J_1 \pm}  M_{J_2 \pm}|
		=
		\left|\frac{b_{J_1}}{2} \pm \frac{b_{J_2}}{2}\right|
		=
		\left|\frac{b_{J_1} \pm b_{J_2}}{2}\right|
		\in\mathbb{N}
		.
	\end{equation}
	In particular, for $H = \{k-1\}$ we obtain $C_H = 2^{2^{k-1}}+1$ and
	\begin{equation}
		b_H =
		\left( 2^{2^{k-1}}+1 - \frac{a}{2^{2^{k-1}}+1} \right)/2
		=
		\left(2^{2^{k-1}}+1 - \left(  2^{2^{k-1}}-1 \right) \right)/2
		=
		1
		.
	\end{equation}
	Then, one of the distances is
	\begin{equation}
		|M_{H+}  M_{H-}|
		=
		\left|\frac{b_{H}}{2} - \frac{-b_{H}}{2}\right|
		=
		\left|\frac{1}{2} - \frac{-1}{2}\right|
		= 1
		.
	\end{equation}

@@

	So, $M = \{ M_{J\pm}, N\}$ is indeed a planar integral point set of $2^k+1$ points,
	and distance 1 occurs in $M$.
	Reminding that $k$ can be taken arbitrary large, we conclude that we can construct
	a planar integral point set of arbitrary large count of points so that the distance 1 occurs in that set.
\end{construction}

@@

	The area of the triangle $M_1 M_3 M_4$ is rational because $|M_3 - M_4| \in \mathbb{Z}$.
	Thus, the set $M$ is of characteristic 1,
	and there is a Cartesian coordinate system such that $M_1=(-1/2,0)$, $M_2=(1/2,0)$, $M_3=(0, a/2)$
	(see~\cite[Theorem 4]{our-vmmsh-2018}).
	It is clear that $a\neq 0$.
	We set $b = |M_1 - M_3|$ and $O=(0,0)$.
	Applying the Pythagorean theorem to triangle $OM_1M_3$, we obtain the following Diophantine equation:
	\begin{equation}
		\frac{1}{4} + \frac{a^2}{4} = b^2
		,
	\end{equation}
	or, which is the same,
	\begin{equation}
		1 + a^2 = (2b)^2
		,
	\end{equation}
	that has no integral solutions.
	This contradiction concludes the proof.
\end{proof}

@@

If the answer for Problem~\ref{prb:square_side_length_and_diameter_linear} is not $\delta = 1$ and $h = 0$,
then the linear estimate can be slightly improved.

However, Problems~\ref{prb:square_side_length_and_diameter_power} and~\ref{prb:square_side_length_and_diameter_linear}
do not seem to show the way for improving the estimate
due to the following well-known theorem,
which can be proved by an application of inversion~\cite{solymosi2010question}.
\begin{theorem}
	\label{thm:rational_set_on_circle}
	There exists a dense subset $P$ of the unit circle such that for any $P_1, P_2 \in P$
	the distance $|P_1 P_2|$ is rational.
\end{theorem}
We conjecture that thorough investigation of existing examples of circular integral point sets
~\cite{bat2018number,harborth1993upper,piepmeyer1996maximum}
will lead to the negative answers for the
Problems~\ref{prb:square_side_length_and_diameter_power} and~\ref{prb:square_side_length_and_diameter_linear}.

@@

\begin{problem}
	\label{prb:minimal_shape}
	What is the minimal shape $\mathcal{S}$ such that any $M\in\mathfrak{M}(2,\mathbb{N})$ with $\operatorname{diam} M = d$
	is situated in  $\mathcal{S}$?
\end{problem}
This problem in another generalization of Lemma~\ref{lem:square_container}.
We know that $\mathcal{S}$ is not a circle of diameter $d$
(the counterexample is a right triangle of side length $d$)
and we conjecture that $\mathcal{S}$ is not a Reuleaux triangle of width $d$
(the probable counterexample belongs to a circle of diameter $d$, see Theorem~\ref{thm:rational_set_on_circle}).

@@

When we talk about integral point sets containing distance 1, another problem arises.

\begin{problem}
	For $M\in\mathfrak{M}(m,n)$, how many times can the distance 1 occur in $M$?
\end{problem}
The answer for $m=2$ is given by Theorem~\ref{thm:minimeter_1_planar}.
For greater $m$, we will be brave enough to put forward the following conjecture.
\begin{conjecture}
	Consider $M\in\mathfrak{M}(m,n)$, $M_1,M_2\in M$, $|M_1 M_2| =1$.
	Then either $n = m+1$ or $M$ is a ``blowup'' of a facher set and distance 1 occurs in $M$
	not more than $1+\frac{(m-1)^2 - (m-1)}{2}$ times.
\end{conjecture}
Such as estimate is based on the quantity of edges in $(m-2)$-dimensional simplex.

>>>>

\section{Introduction}
Let $\mathbb{N}$ be the set of all positive integers and let $|M_1 M_2|$ denote the Euclidean distance
between points $M_1$ and $M_2$ in a finite-dimensional space $\mathbb{R}^m$
(and, more generally, let $|\Delta|$ denote the length of line segment $\Delta$).
An \textit{integral point set} in $m$-dimensional Euclidean space is a point set $M$ such that all the distances between the
points of $M$ are integers and $M$ is not situated on an $(m-1)$-dimensional hyperplane.
Erd\"os and Anning proved~\cite{anning1945integral,erdos1945integral} that every integral point set consists of a finite number of points.
Taking the fact into account, we will denote the set of all integral point sets of $n$ points in $m$-dimensional Euclidean space with
$\mathfrak{M}(m,n)$ (using the notation in~\cite{our-vmmsh-2018})
and denote the set of all integral point sets in $m$-dimensional Euclidean space with $\mathfrak{M}(m,\mathbb{N})$.
The symbol $\# M$ will be used for cardinality of $M$, i.e. the quantity of points in $M$ in our case.

@@

In this section,
we improve the lower bound for minimum diameter of planar integral point sets by
involving Point Packing in a Square problem.
Below we introduce the problem, basic notions and results.

@@

\begin{definition}
	A \textit{cross} for points $M_1$ and $M_2$, denoted by $cr(M_1,M_2)$, is the union of two straight lines:
	the line which the points $M_1$ and $M_2$ are situated on,
	and the bisector of line segment $M_1 M_2$.
\end{definition}

@@

Performing some simple manipulations with the upper bound in Theorem~\ref{thm:varphi_k_bounds},
one can derive the following proposition.
\begin{proposition}
	\label{varphi_n_where_n_geq_21492}
	For $n \geq 21492$ we have
	\begin{equation}
		\varphi_n \leq \varphi_{n-1} \leq \frac{\beta}{\sqrt{n-2}}
		,
	\end{equation}
	where
	\begin{equation}
		\beta = \frac{1}{\sqrt{21490}} + \sqrt{ \frac{2}{\sqrt{3}} + \frac{1}{21490} } < 1.07464
		.
	\end{equation}
\end{proposition}

Now we need to estimate the cardinality of an intersection of an integral point set with a line segment.
There is the following estimate for minimum diameter of planar integral point sets with many collinear points.
\begin{theorem}~\cite[Theorem 4]{kurz2008minimum}
	For $\delta > 0$, $\varepsilon > 0$, and $P\in\mathfrak{M}(2,n)$ with
	at least $n^\delta$ collinear points there exists a $n_0 (\varepsilon)$
	such that for all $n \geq n_0 (\varepsilon)$ we have
	\begin{equation}
		\operatorname{diam} P \geq n^{\frac{\delta}{4 \log 2(1+\varepsilon)}\log \log n}
		.
	\end{equation}
\end{theorem}
However, the estimate is rather unsuitable for our needs,
as it does not provide the values of all the constants.
To obtain the desired estimate, we will now prove a generalization of~\cite[lemma 3]{our-vmmsh-2018}.

@@

\begin{proposition}
	\label{obs:rho_curves}
	If points $M_1,M_2,M_3,M_4$ are situated on a straight line,
	and equality $|M_1 M_2| = |M_3 M_4|$ holds but line segments $M_1 M_2$ and $M_3 M_4$ do not coincide,
	then for a fixed $k$ the $\rho(k,M_1 M_2)$-curve and the $\rho(k,M_3 M_4)$-curve
	do not intersect.
\end{proposition}

@@

\begin{proof}
	Let $f(k)$ denote the mininal length of a line segment
	that shares $k$ points with an integral point set.
	We observe that $f(k) > f (k-1)$.
	Due to Lemma~\ref{lem:line_segment_with n_squared_plus_one_points},
	$f(n^2+1) \geq \frac{2}{3}n^3+\frac{1}{2}n^2-\frac{1}{6}n$.

@@

\begin{theorem}
	\label{thm:main_estimate}
	If $n\geq 4$, then $d(2,n) \geq \gamma (n - 2)$,
	where
	\begin{multline}
		0.46530... =
		3^{1/4} \cdot 2^{-3/2} >
		\\ >
		\gamma = \frac{\sqrt{16 {{\left( \sqrt{\frac{2}{\sqrt{3}}+\frac{1}{21490}}+\frac{1}{\sqrt{21490}}\right) }^{2}}+\frac{14791716}{{{\left( 2593 \sqrt{647}-5823\right) }^{2}}}}-\frac{3846}{2593 \sqrt{647}-5823}}{8 {{\left( \sqrt{\frac{2}{\sqrt{3}}+\frac{1}{21490}}+\frac{1}{\sqrt{21490}}\right) }^{2}}}
		>\\
		> 0.45557
		> \frac{5}{11}
		.
	\end{multline}
\end{theorem}

@@

Facher sets are the simplest ones of planar integral point sets.
It is shown that for $9\leq n \leq 122$ all the optimal sets are facher~\cite{kurz2008minimum}.
For every cardinality $n$ and every squarefree number $q$
there exists a facher set $M\in\mathfrak{M}(2,n)$ with characteristic $q$~\cite[Theorem 5]{our-vmmsh-2018}.
In~\cite{antonov2008maximal}, the facher sets of characteristic 1 were investigated; they were called \textit{semi-crabs}.

@@

	We will now define the coordinates of the points as following:
	\begin{equation}
		M_{J\pm} =\left(\pm\frac{b_J}{2}, 0\right)
		,
		~~
		N   =\left(0, \frac{\sqrt{a}}{2}\right)
		.
	\end{equation}
	Then the distances are:
	\begin{multline}
		|N M_{J\pm}|
		=
		\left(\frac{b_J^2}{4} + \frac{a}{4}\right)^{1/2}
		=
		\frac{1}{2}\left(\left(\frac{c_J - a/c_J}{2}\right)^2 + a\right)^{1/2}
		=
		\\=
		\frac{1}{2}\left( \left(\frac{c_J}{2}\right)^2 - \frac{a}{2} + \left(\frac{a}{c_j}\right)^2 + a\right)^{1/2}
		=
		\frac{1}{2}\left( \left(\frac{c_J}{2}\right)^2 + \frac{a}{2} + \left(\frac{a}{c_j}\right)^2    \right)^{1/2}
		=
		\\=
		\frac{1}{2}\left(\left(\frac{c_J + a/c_J}{2}\right)^2\right)^{1/2}
		=
		\frac{1}{2}\left(\frac{c_J + a/c_J}{2}\right)
		=
		\frac{g_J}{2}
		\in\mathbb{N}
		,
	\end{multline}
	\begin{equation}
		|M_{J_1 \pm}  M_{J_2 \pm}|
		=
		\left|\frac{b_{J_1}}{2} \pm \frac{b_{J_2}}{2}\right|
		=
		\left|\frac{b_{J_1} \pm b_{J_2}}{2}\right|
		\in\mathbb{N}
		.
	\end{equation}
	In particular, for $H = \{k-1\}$ we obtain $C_H = 2^{2^{k-1}}+1$ and
	\begin{equation}
		b_H =
		\left( 2^{2^{k-1}}+1 - \frac{a}{2^{2^{k-1}}+1} \right)/2
		=
		\left(2^{2^{k-1}}+1 - \left(  2^{2^{k-1}}-1 \right) \right)/2
		=
		1
		.
	\end{equation}
	Thus, one of the distances is
	\begin{equation}
		|M_{H+}  M_{H-}|
		=
		\left|\frac{b_{H}}{2} - \frac{-b_{H}}{2}\right|
		=
		\left|\frac{1}{2} - \frac{-1}{2}\right|
		= 1
		.
	\end{equation}

@@

	So, $M = \{ M_{J\pm}, N\}$ is indeed a planar integral point set of $2^k+1$ points,
	and distance 1 occurs in $M$.
	Reminding that $k$ can be taken arbitrary large, we conclude that we can construct
	a planar integral point set of arbitrary large cardinality so that the distance 1 occurs in that set.
\end{construction}

@@

	The area of the triangle $M_1 M_3 M_4$ is rational because $|M_3 - M_4| \in \mathbb{Z}$.
	Thus, the characteristic of $M$ is 1,
	and there is a Cartesian coordinate system such that $M_1=(-1/2,0)$, $M_2=(1/2,0)$, $M_3=(0, a/2)$
	(see~\cite[Theorem 4]{our-vmmsh-2018}).
	It is clear that $a\neq 0$.
	We set $b = |M_1 - M_3|$ and $O=(0,0)$.
	Applying the Pythagorean theorem to triangle $OM_1M_3$, we obtain the following Diophantine equation:
	\begin{equation}
		\frac{1}{4} + \frac{a^2}{4} = b^2
		,
	\end{equation}
	or, which is the same,
	\begin{equation}
		1 + a^2 = (2b)^2
		,
	\end{equation}
	that has no integral solutions.
	This contradiction concludes the proof.
\end{proof}

@@

If the answer for Problem~\ref{prb:square_side_length_and_diameter_linear} is not $\delta = 1$ and $h = 0$,
then the linear bound can be slightly improved.

However, Problems~\ref{prb:square_side_length_and_diameter_power} and~\ref{prb:square_side_length_and_diameter_linear}
do not seem to show the way for improving the bound
due to the following well-known theorem,
which can be proved by an application of inversion~\cite{solymosi2010question}.
\begin{theorem}
	\label{thm:rational_set_on_circle}
	There exists a dense subset $P$ of the unit circle such that for any $P_1, P_2 \in P$
	the distance $|P_1 P_2|$ is rational.
\end{theorem}
We conjecture that thorough investigation of existing examples of circular integral point sets
~\cite{bat2018number,harborth1993upper,piepmeyer1996maximum}
will lead to the negative answers for the
Problems~\ref{prb:square_side_length_and_diameter_power} and~\ref{prb:square_side_length_and_diameter_linear}.

@@

\begin{problem}
	\label{prb:minimal_shape}
	What is the minimal shape $\mathcal{S}$ such that any $M\in\mathfrak{M}(2,\mathbb{N})$ with $\operatorname{diam} M = d$
	is situated in  $\mathcal{S}$?
\end{problem}
This problem is another generalization of Lemma~\ref{lem:square_container}.
We know that $\mathcal{S}$ is not a circle of diameter $d$
(the counterexample is a right triangle of side length $d$)
and we conjecture that $\mathcal{S}$ is not a Reuleaux triangle of width $d$
(the probable counterexample belongs to a circle of diameter $d$, see Theorem~\ref{thm:rational_set_on_circle}).

@@

For planar integral point sets containing distance 1, we can put forward the following problem.
\begin{problem}
	For given $n\geq 3$ and squarefree number $q \neq 1$,
	is there $M\in\mathfrak{M}(2,n)$, such that characteristic of $M$ is $q$
	and distance 1 occurs in $M$?
\end{problem}

When we talk about integral point sets containing distance 1 of higher dimensions, another problem arises.

\begin{problem}
	For $M\in\mathfrak{M}(m,n)$, how many times can the distance 1 occur in $M$?
\end{problem}
The answer for $m=2$ is given by Theorem~\ref{thm:minimeter_1_planar}.
For greater $m$, we will be brave enough to put forward the following conjecture.
\begin{conjecture}
	Consider $M\in\mathfrak{M}(m,n)$, $M_1,M_2\in M$, $|M_1 M_2| =1$.
	Then either $n = m+1$ or $M$ is a ``blowup'' of a facher set and distance 1 occurs in $M$
	not more than $1+\frac{(m-1)^2 - (m-1)}{2}$ times.
\end{conjecture}
Such as estimate is based on the quantity of edges in an $(m-2)$-dimensional simplex.
